Interspecies comparisons of Mg/Ca ratios in limpet shells

### 2024-06-23 Manuscript.qmd
Intro

While recent studies of particularly *Patella* sp. in the Mediterranean and Southwest Europe have provided promising results [@Hausmann2019-fi; @Garcia-Escarzaga2015-jc; @Garcia-Escarzaga2018-nf].

@@

@@

| Species            | Locality       | Specimen | Correlation R^2^        | Study                                                                                                                              |
|-------------|-------------|-------------|-------------|----------------------|
| *Patella depressa* | Northern Spain | LAN541   | 0.87                    | [**10.3390/app11072959**](https://doi.org/10.3390/app11072959)                                                                     |
|                    |                | LAN545   | 0.86                    |                                                                                                                                    |
|                    |                | LAN554   | 0.78                    |                                                                                                                                    |
|                    |                | LAN559   | 0.82                    |                                                                                                                                    |
| *Patella caerulea* | Croatia        | ISTPC1   | 0.9                     | 10.1038/s41598-019-39959-9                                                                                                         |
|                    |                | ISTPC2   | 0.84                    |                                                                                                                                    |
|                    | Crete          | AF1911A  | 0.91[^1]                |                                                                                                                                    |
|                    |                | AF3003A  | 0.92[^2]                |                                                                                                                                    |
|                    | Israel         | AKKPC2   | 0.96                    |                                                                                                                                    |
|                    |                | AKKPC3   | 0.89                    |                                                                                                                                    |
|                    |                | FRMPC1   | 0.84                    |                                                                                                                                    |
|                    |                | FRMPC2   | 0.96                    |                                                                                                                                    |
|                    | Libya          | MO31A    | 0.83                    |                                                                                                                                    |
|                    |                | MP64A    | 0.33                    |                                                                                                                                    |
|                    |                | MP67A    | 0.96                    |                                                                                                                                    |
|                    |                | MP68A    | 0.81                    |                                                                                                                                    |
|                    | Malta          | MA10     | 0.82                    |                                                                                                                                    |
|                    | Tunisia        | TUNPC1   | 0.81                    |                                                                                                                                    |
|                    |                | TUNPC2   | 0.78                    |                                                                                                                                    |
|                    | Turkey         | ANTPC1   | 0.95                    |                                                                                                                                    |
|                    |                | ANTPC2   | 0.93                    |                                                                                                                                    |
|                    |                | KIZPC1   | 0.94                    |                                                                                                                                    |
|                    |                | KIZPC2   | 0.86                    |                                                                                                                                    |
| Patella rustica    | Gibraltar      | JL1      | 0.02                    | [doi.org/10.1016/j.epsl.2011.05.054](https://doi.org/10.1016/j.epsl.2011.05.054 "Persistent link using digital object identifier") |
|                    |                | JL2      | 0.8 (0.79)              |                                                                                                                                    |
| Patella caerulea   | Gibraltar      | JM00     | 0.69 (0.79)             |                                                                                                                                    |
|                    |                | JM30     | 0.83 (0.79)             |                                                                                                                                    |
| Patella vulgata    | Orkney         | ORK-LT5  | not reported, here 0.88 | doi.org/10.1016/j.palaeo.2016.10.021 and this study                                                                                |
|                    |                |          |                         |                                                                                                                                    |
|                    |                |          |                         |                                                                                                                                    |
|                    |                |          |                         |                                                                                                                                    |

: Overview of comparative correlations {#tab:correlations}

>>>>

Limpet shells are commonly found within archaeological sites and and past shorelines due to their robust carbonate structure and the long-term use of limpets as a marine food source. They have successfully been studied in the past in the context of coastal subsistence economies, site occupation on seasonal [@shackleton1973; @Parker2018-wf; @Bosch2018-ud] or long-term scale [@Ortiz2015-mr], as well as palaeotemperature [@Fenger2007-gf; @Surge2012-ba; @Wang2012-ee; @Colonese2012-ct; @Ferguson2011-zl]. Determining past sea surface temperature (SST) change largely relies on the measurement of δ^18^O-values within the calcitic parts of the shell, but attempts have been made to also use elemental ratios, such as magnesium to calcium (Mg/Ca), to have an alternative measure, that potentially provides a more accurate SST estimate than δ^18^O-values, which are also affected by changes in salinity. In addition, the data acquisition of elemental ratios — either through laser-ablation-isotope-ratio-mass-spectrometry (LA-ICP-MS) or laser induced breakdown spectroscopy (LIBS), can be much faster and cost-effective, increasing the number of specimens that can be studied overall [@Durham2017-fh; @Hausmann2023-ih].

Problematically, links between Mg/Ca ratios and SST changes in limpets have also been shown to be unreliable [@Graniero2015-zv] and even where they were successful, anomalous patterns in some specimens had to be filtered out, reducing the overall robustness of the results of those successful studies [@Ferguson2011-zl].

While recent research particularly of *Patella* sp. in the Mediterranean and Southwest Europe have provided promising results [@Hausmann2019-fi; @Garcia-Escarzaga2015-jc; @Garcia-Escarzaga2018-nf]. There remains unclarity for Atlantic limpet species, particularly since it was last shown here, that they are not reliable recorders of palaeotemperature [@Graniero2015-zv].

@@

The analysed specimens, their origin and studies with background information can be found in Table 1.

| Context        | Study              | Species          | Location                                          | Sample ID     | Previous analyses                                           |
|------------|------------|------------|------------|------------|------------|
| Modern         | [@Nicastro2020-ih] | *N. deaureata*   | Bahía Cambaceres Exterior - Tierra del Fuego (AR) | ND-1016-3     | Stable oxygen and carbon isotope analysis                   |
|                |                    |                  |                                                   | ND-1016-4     |                                                             |
|                |                    | *N. magellanica* |                                                   | NM-1016-1     |                                                             |
|                |                    |                  |                                                   | NM-1016-3     |                                                             |
|                | [@Graniero2017-io] | *P. vulgata*     | Rack Wick Bay - Westray (UK)                      | ORK-LT5       | Stable oxygen and carbon isotope analysis; Mg, Li, Sr, Ca,. |
| Archaeological | [@Surge2012-ba]    | *P. vulgata*     | Rack Wick Bay - Westray (UK)                      | QG1-7188-1    | Stable oxygen and carbon isotope analysis                   |
|                |                    |                  |                                                   | QG1-7189-2    |                                                             |
|                |                    |                  |                                                   | QG2-1061-1    |                                                             |
|                |                    |                  |                                                   | QG2-1064-1    |                                                             |
|                |                    |                  |                                                   | QG2-7180-1    |                                                             |
|                |                    |                  |                                                   | QG2-7180-2    |                                                             |
|                |                    |                  |                                                   | QG1-7246-1    |                                                             |
|                | This study         | *N. magellanica* | Heshkaia 35                                       | Heshkaia_35-1 | *none*                                                      |
|                |                    | *N. deaureata*   |                                                   | Heshkaia_35-2 |                                                             |
|                |                    | *N. deaureata*   |                                                   | Heshkaia_35-3 |                                                             |
|                |                    | *N. magellanica* |                                                   | Heshkaia_35-4 |                                                             |
|                |                    | *N. magellanica* |                                                   | Heshkaia_35-5 |                                                             |
|                |                    | *N. magellanica* |                                                   | Heshkaia_35-6 |                                                             |
|                |                    | *N. magellanica* |                                                   | Heshkaia_35-7 |                                                             |
|                |                    | *N. deaureata*   |                                                   | Heshkaia_35-8 |                                                             |
|                |                    | *N. magellanica* |                                                   | Heshkaia_35-9 |                                                             |

: Table 1 Overview of the modern and archaeological limpet specimens analysed in this study {#tab:specimens}

@@

| Species       | Locality       | Specimen | Correlation R^2^        | Study                             |
|---------------|---------------|---------------|---------------|---------------|
| *P. depressa* | Northern Spain | LAN541   | 0.87                    | [@garcía-escárzaga2021]           |
|               |                | LAN545   | 0.86                    |                                   |
|               |                | LAN554   | 0.78                    |                                   |
|               |                | LAN559   | 0.82                    |                                   |
| *P. caerulea* | Croatia        | ISTPC1   | 0.9                     | [@Hausmann2019-fi]                |
|               |                | ISTPC2   | 0.84                    |                                   |
|               | Crete          | AF1911A  | 0.91[^1]                |                                   |
|               |                | AF3003A  | 0.92[^2]                |                                   |
|               | Israel         | AKKPC2   | 0.96                    |                                   |
|               |                | AKKPC3   | 0.89                    |                                   |
|               |                | FRMPC1   | 0.84                    |                                   |
|               |                | FRMPC2   | 0.96                    |                                   |
|               | Libya          | MO31A    | 0.83                    |                                   |
|               |                | MP64A    | 0.33                    |                                   |
|               |                | MP67A    | 0.96                    |                                   |
|               |                | MP68A    | 0.81                    |                                   |
|               | Malta          | MA10     | 0.82                    |                                   |
|               | Tunisia        | TUNPC1   | 0.81                    |                                   |
|               |                | TUNPC2   | 0.78                    |                                   |
|               | Turkey         | ANTPC1   | 0.95                    |                                   |
|               |                | ANTPC2   | 0.93                    |                                   |
|               |                | KIZPC1   | 0.94                    |                                   |
|               |                | KIZPC2   | 0.86                    |                                   |
| *P. rustica*  | Gibraltar      | JL1      | 0.02                    | [@Ferguson2011-zl]                |
|               |                | JL2      | 0.8 (0.79)              |                                   |
| *P. caerulea* | Gibraltar      | JM00     | 0.69 (0.79)             |                                   |
|               |                | JM30     | 0.83 (0.79)             |                                   |
| *P. vulgata*  | Orkney         | ORK-LT5  | not reported, here 0.88 | [@Graniero2017-io] and this study |
|               |                |          |                         |                                   |
|               |                |          |                         |                                   |
|               |                |          |                         |                                   |

: Table 2: Overview of comparative correlations {#tab:correlations}

### 2024-06-23 Manuscript.tex
Intro

\begin{abstract}
This document is a template demonstrating the Aps format.
\end{abstract}

\begin{keyword}
    keyword1 \sep 
    keyword2
\end{keyword}
\end{frontmatter}

{\centering \includegraphics{Manuscript_files/figure-pdf/Raw Data-1.pdf}

\caption{Raw Mg/Ca and δ\textsuperscript{18}O Data}

@@

>>>>

\begin{abstract}
This study provides a short reassessment of the use of Magnesium to
Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea
surface temperatures. While \emph{Patella vulgata} along the Spanish
shoreline has since then repeatedly produced reliable correlations
between sea surface temperature and Mg/Ca ratios, this is not the case
for other patelloid species. \emph{Patella vulgata} and \emph{Nacella
deaureata} have been studied using Mg/Ca with mixed or contrary results.
In this study, we present elemental maps of various such species
together with stable oxygen isotope values for some of the specimens.
Our dataset also includes specimens that were previously unsuccessful in
providing significant correlations between δ\textsuperscript{18}O and
Mg/Ca ratios. By reassessing these previous specimens and including a
wider range of modern and archaeological samples from three patelloid
species (\emph{P. vulgata}, \emph{N. deaureata}, and \emph{N.
magellanica}) we further add to the growing set of evidence for the
reliable use of Mg/Ca ratios to detect palaeotemperature change and
serve as a means to determine ontogenetic age and season of capture as
well as to reveal locations of interest within the growth record
(i.e.~annual temperature minima and maxima) for the targeted analysis
using δ\textsuperscript{18}O or clumped oxygen isotope analysis.
\end{abstract}

\begin{keyword}
    Sclerochronology \sep Limpets \sep Elemental Ratio \sep 
    Mg/Ca
\end{keyword}
\end{frontmatter}

\section{Introduction}\label{Introduction}

Limpet shells are commonly found within archaeological sites and and
past shorelines due to their robust carbonate structure and the
long-term use of limpets as a marine food source. They have successfully
been studied in the past in the context of coastal subsistence
economies, site occupation on seasonal
\citep{shackleton1973, Parker2018-wf, Bosch2018-ud} or long-term scale
\citep{Ortiz2015-mr}, as well as palaeotemperature
\citep{Fenger2007-gf, Surge2012-ba, Wang2012-ee, Colonese2012-ct, Ferguson2011-zl}.
Determining past sea surface temperature (SST) change largely relies on
the measurement of δ\textsuperscript{18}O-values within the calcitic
parts of the shell, but attempts have been made to also use elemental
ratios, such as magnesium to calcium (Mg/Ca), to have an alternative
measure, that potentially provides a more accurate SST estimate than
δ\textsuperscript{18}O-values, which are also affected by changes in
salinity. In addition, the data acquisition of elemental ratios ---
either through laser-ablation-isotope-ratio-mass-spectrometry
(LA-ICP-MS) or laser induced breakdown spectroscopy (LIBS), can be much
faster and cost-effective, increasing the number of specimens that can
be studied overall \citep{Durham2017-fh, Hausmann2023-ih}.

Problematically, links between Mg/Ca ratios and SST changes in limpets
have also been shown to be unreliable \citep{Graniero2015-zv} and even
where they were successful, anomalous patterns in some specimens had to
be filtered out, reducing the overall robustness of the results of those
successful studies \citep{Ferguson2011-zl}.

While recent research particularly of \emph{Patella} sp. in the
Mediterranean and Southwest Europe have provided promising results
\citep{Hausmann2019-fi, Garcia-Escarzaga2015-jc, Garcia-Escarzaga2018-nf}.
There remains unclarity for Atlantic limpet species, particularly since
it was last shown here, that they are not reliable recorders of
palaeotemperature \citep{Graniero2015-zv}.

\section{Materials and Methods}\label{Methods}

\subsection{Limpet specimens}\label{limpet-specimens}

\subsubsection{Modern specimens}\label{modern-specimens}

The analysed specimens, their origin and studies with background
information can be found in Table 1.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}@{}}
\caption{Table 1 Overview of the modern and archaeological limpet
specimens analysed in this study \{\#tab:specimens\}}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Context
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Study
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Species
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Location
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Sample ID
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Previous analyses
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Context
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Study
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Species
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Location
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Sample ID
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Previous analyses
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Modern & \citep{Nicastro2020-ih} & \emph{N. deaureata} & Bahía
Cambaceres Exterior - Tierra del Fuego (AR) & ND-1016-3 & Stable oxygen
and carbon isotope analysis \\
& & & & ND-1016-4 & \\
& & \emph{N. magellanica} & & NM-1016-1 & \\
& & & & NM-1016-3 & \\
& \citep{Graniero2017-io} & \emph{P. vulgata} & Rack Wick Bay - Westray
(UK) & ORK-LT5 & Stable oxygen and carbon isotope analysis; Mg, Li, Sr,
Ca,. \\
Archaeological & \citep{Surge2012-ba} & \emph{P. vulgata} & Rack Wick
Bay - Westray (UK) & QG1-7188-1 & Stable oxygen and carbon isotope
analysis \\
& & & & QG1-7189-2 & \\
& & & & QG2-1061-1 & \\
& & & & QG2-1064-1 & \\
& & & & QG2-7180-1 & \\
& & & & QG2-7180-2 & \\
& & & & QG1-7246-1 & \\
& This study & \emph{N. magellanica} & Heshkaia 35 & Heshkaia\_35-1 &
\emph{none} \\
& & \emph{N. deaureata} & & Heshkaia\_35-2 & \\
& & \emph{N. deaureata} & & Heshkaia\_35-3 & \\
& & \emph{N. magellanica} & & Heshkaia\_35-4 & \\
& & \emph{N. magellanica} & & Heshkaia\_35-5 & \\
& & \emph{N. magellanica} & & Heshkaia\_35-6 & \\
& & \emph{N. magellanica} & & Heshkaia\_35-7 & \\
& & \emph{N. deaureata} & & Heshkaia\_35-8 & \\
& & \emph{N. magellanica} & & Heshkaia\_35-9 & \\
\end{longtable}

\subsubsection{Archaeological specimens}\label{archaeological-specimens}

\subsection{Oxygen isotopes}\label{oxygen-isotopes}

\subsection{Mg/Ca ratios}\label{mgca-ratios}

\section{Results}\label{Results}

\subsection{Patella vulgata}\label{patella-vulgata}

\subsection{Nacella deaureata}\label{nacella-deaureata}

\subsection{Nacella magellanica}\label{nacella-magellanica}

\section{}\label{section}

@@

\section{Discussion}\label{discussion}

\subsection{Other correlations}\label{other-correlations}

{\centering \includegraphics{Manuscript_files/figure-pdf/Ferguson Data-1.pdf}

\caption{Correlation graphs for Ferguson et al.~specimens}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}@{}}
\caption{Table 2: Overview of comparative correlations
\{\#tab:correlations\}}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Species
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Locality
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Specimen
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Correlation R\textsuperscript{2}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Study
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Species
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Locality
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Specimen
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Correlation R\textsuperscript{2}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Study
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\emph{P. depressa} & Northern Spain & LAN541 & 0.87 &
\citep{garcía-escárzaga2021} \\
& & LAN545 & 0.86 & \\
& & LAN554 & 0.78 & \\
& & LAN559 & 0.82 & \\
\emph{P. caerulea} & Croatia & ISTPC1 & 0.9 & \citep{Hausmann2019-fi} \\
& & ISTPC2 & 0.84 & \\
& Crete & AF1911A & 0.91\footnote{SST only, no other geochemical data
  available} & \\
& & AF3003A & 0.92\footnote{SST only, no geochemical data available}
& \\
& Israel & AKKPC2 & 0.96 & \\
& & AKKPC3 & 0.89 & \\
& & FRMPC1 & 0.84 & \\
& & FRMPC2 & 0.96 & \\
& Libya & MO31A & 0.83 & \\
& & MP64A & 0.33 & \\
& & MP67A & 0.96 & \\
& & MP68A & 0.81 & \\
& Malta & MA10 & 0.82 & \\
& Tunisia & TUNPC1 & 0.81 & \\
& & TUNPC2 & 0.78 & \\
& Turkey & ANTPC1 & 0.95 & \\
& & ANTPC2 & 0.93 & \\
& & KIZPC1 & 0.94 & \\
& & KIZPC2 & 0.86 & \\
\emph{P. rustica} & Gibraltar & JL1 & 0.02 & \citep{Ferguson2011-zl} \\
& & JL2 & 0.8 (0.79) & \\
\emph{P. caerulea} & Gibraltar & JM00 & 0.69 (0.79) & \\
& & JM30 & 0.83 (0.79) & \\
\emph{P. vulgata} & Orkney & ORK-LT5 & not reported, here 0.88 &
\citep{Graniero2017-io} and this study \\
& & & & \\
& & & & \\
& & & & \\
\end{longtable}

\subsection{Comparison of ORK-LT5}\label{comparison-of-ork-lt5}

\includegraphics{Manuscript_files/figure-pdf/unnamed-chunk-1-1.pdf}

\includegraphics{Manuscript_files/figure-pdf/subsample-1.pdf}
